Nucleotide Diversity Loss on a Plant Y Chromosome Following Recent Recombination Suppression

### 2016-11-13 manuscript/manuscript.tex
final edits up to results, added new ms version for Spencer and Stephen

\subsection*{Neutral predictions and the effect of sex ratio bias on diversity}
To test whether the levels of diversity on X, Y and autosomal chromosomes could be explained by neutral processes, we compared our diversity estimates to predictions from a model that incorporates variance of progeny number for males and females, and gives the predicted $N_{e}$ for males and females when this variance is either smaller or larger than Poisson \citep{kimura1964number, hedrick2011genetics}. As we were primarily interested in the possible influence of variance in male reproductive success on Y-linked $N_{e}$, and the parameter space within which such a model might generate Y/A and X/A ratios consistent with our data, we ignored non-poission variance in female offspring, and calculated the male $N_{e}$ as a function of variance in offspring number as:

@@

where $k_{m}$ and $V_{k_{m}}$ are the mean and variance in male offspring number, respectively, in a population of $N_{m}$ reproducing males and $\overline{k}$ is the total mean number of progeny of males and females \citep{kimura1964number}. Given the expected values for $N_e_{m}$ across a range of values for $V_{k_{m}}$ and $N_{m}$, the corresponding expectation for the $N_{e}$ of autosomes and sex chromosomes at equilibrium were calculated as:

@@

following \citep{wright1931evolution}. Note that with equal sex ratios, and Poisson distributed offspring numbers ($V_{k_{m}} = \overline{k}$), then $N_{e_{X}}/N_{e_{A}} = 0.75$, and $N_{e_{Y}}/N_{e_{A}} = 0.25$. Figure 1 shows these predictions as a function of the sex ratio for a range values of $V_{k_{m}}$, and we tested the fit of our empirical diversity ratios to these predictions (shown in Figure 2) assuming the estimated \textit{R. hastatulus} sex ratio of 0.6 \citep{pickup2013influence}.

@@

\subsection*{Simulations of purifying selection}
To test whether our observed level of Y-chromosome diversity could be explained by the effects of linked purifying selection, we followed the approach used in \citep{Wilsonsayres2014} and conducted forward-time simulations of a non-recombining Y chromosome using the software SFSCODE \citep{hernandez2008flexible} and compared our empirical diversity estimates on the Y with those from simulations. This approach was implemented rather than using analytical predictions of background selection for a few reasons. First,... because the equilibrium background selection model over-predicts the reduction in diversity when there are many linked sites under selection \citep{KaiserCharlesworth}, as is expected to be the case for large Y chromosomes that lack crossing over.

To parameterize the forward simulations, we estimated the distribution of fitness effects of deleterious mutations (DFE) using our polymorphism data on the X-chromosome with the program DFE-alpha (\X Keightley and Eyre-Walker ref). This program fits a gamma distribution of selection coefficients to the observed frequency distribution of nonsynonymous and synonymous polymorphism from the population genetic analyses above. We used this estimated gamma distribution in our forward simulations. To examine the expected reduction in diversity, we varied the number of sites under selection, and obtained an approximate likelihood surface for our observed data based on the proportion of simulations that matched the observed estimate of $\pi_{synonymous}$ . To make the simulation output comparable to our data, we initialized the simulations with our empirically estimated autosomal $\theta$, adjusted for the expected neutral reduction in effective population size on the Y chromosome under a sex ratio of $N_{f}/(N_{f}+N_{m})=0.6$, and we sampled 6 haploid chromosomes per simulation. Simulated sequences contained 45,331 of linked neutral sequence from which we calculated diversity, matching the number of synonymous sites sampled in our empirical analyses on the Y chromosome (see Supporting Information).

@@

>>>>

\subsection*{Neutral predictions and the effect of sex ratio bias on diversity}
To test whether our estimated levels of diversity on X, Y and autosomal chromosomes could be explained by neutral processes, we compared our estimated levels diversity to predictions from a model of the expected $N_{e}$ of males and females when the variance in offpring number is either smaller or larger than Poisson \citep{kimura1964number, hedrick2011genetics}. As we were primarily interested in the possible influence of variance in male reproductive success on Y-linked $N_{e}$, and the parameter space within which such a model might generate Y/A and X/A ratios consistent with our data, we ignored non-poission variance in female offspring, and calculated the male $N_{e}$ as:

@@

where $k_{m}$ and $V_{k_{m}}$ are the mean and variance in male offspring number, respectively, in a population of $N_{m}$ reproducing males, and $\overline{k}$ is the total mean number of progeny \citep{kimura1964number}. Given expected values of $N_e_{m}$ for a range of values for $V_{k_{m}}$ and $N_{m}$, the corresponding expectation for the $N_{e}$ of autosomes and sex chromosomes at equilibrium were calculated as:

@@

following \citep{wright1931evolution}. Note that with equal sex ratios $N_{f} = N_{m}$, and Poisson-distributed offspring numbers ($V_{k} = \overline{k}$), then $N_{e_{X}}/N_{e_{A}} = 0.75$, and $N_{e_{Y}}/N_{e_{A}} = 0.25$. Figure 1 shows these predictions as a function of the sex ratio for a range values of $V_{k_{m}}$, and we tested the fit of our empirical diversity ratios to these predictions (shown in Figure 2) assuming the estimated \textit{R. hastatulus} sex ratio of 0.6 \citep{pickup2013influence}.

@@

\subsection*{Simulations of purifying selection}
To study the effects of purifying selection on expected levels of Y-chromosome diversity, we used a similar approach to \citep{Wilsonsayres2014} and conducted forward-time simulations of haploid Y chromosomes using the software SFSCODE \citep{hernandez2008flexible}. We first estimated the distribution of fitness effects of deleterious mutations from our polymorphism data for X-linked genes using the method of \citep{keightley2007joint}, which fits a gamma distribution of selection coefficients to the observed frequency distribution of nonsynonymous and synonymous polymorphism. We used this estimated gamma distribution to parameterize the simulations, initializing each with our estimated $\theta$ from autosomal genes, but appropriately adjusted to reflect the expected reduction in Y-linked $N_{e}$ for a sex ratio of $N_{f}/(N_{f}+N_{m})=0.6$. To match our sample size and the number of synonymous sites sample from our data (see Supporting Information), the simulations sampled 6 haploid chromosomes, and simulated sequences contained 45,331bp of linked neutral sequence from which we calculated diversity. To examine the expected reduction in diversity relative to neutrality ($\pi/\pi_{0}$) as a function of the number of selected sites ($L$), and to estimate this parameter from our data, we ran simulations over a range of values of $L$ (up to a maximum of $5x10^{6}$) (Figure 3A). For each simulation, we calculated the approximate likelihood of our observed data based on the proportion of simulations in which synonymous diversity, $\pi_{s}_{simulated}$, matched our empirical estimate, $\pi_{s}_{observed}$ (Figure 3B).

@@

This approach was implemented rather than using analytical predictions of background selection for a few reasons. First,... because the equilibrium background selection model over-predicts the reduction in diversity when there are many linked sites under selection \citep{KaiserCharlesworth}, as is expected to be the case for large Y chromosomes that lack crossing over.
